Type theory in type theory with single substitutions

### 2024-09-15 types_2024_postproc/p.tex
stuff

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
This concludes the first-order single substitution semantics of the
SOGAT \ref{eq:tt}. The theory can be extended with new type and term
formers in the same fashion: each rule has to be indexed over
arbitrary contexts and come equipped with a substitution rule. There
is no need to add more structural (substitution calculus) rules.

@@

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we exchange
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
With this we derived an even more minimalistic first-order theory for
the SOGAT (\ref{eq:tt}), essentially relying on the universe and $\Pi$
types. It is clear from the construction that the resulting theory is
isomorphic to the one presented in Section \ref{sec:tt} (and Appendix
\ref{app:tt}), all equations are interderivable. For reference, we
list all the rules of the minimised calculus in Appendix
\ref{app:minimisation}.

@@

We list the rules of the theory derived in Section
\ref{sec:minimisation}.
\begin{alignat*}{10}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\N\ra\Set \\
  & \diamond && : \Con \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
  & \Sub && : \Con\ra\Con\ra\Set \\
  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & [\p][^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_e b[\gamma][\p] \\
  & \q[^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra \q[\gamma^+] =_e \q \\
  & [\p][\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra b[\p ][\langle a\rangle] =_e b \\
  & \q[\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra \q[\langle a\rangle] =_e a \\
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[]' && : (e:B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
  & && \hphantom{{}:{}} (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}

>>>>

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
This concludes the first-order single substitution semantics of the
SOGAT \ref{eq:tt}. For reference, we list the same rules again in
Appendix \ref{app:tt}. The theory can be extended with new type and
term formers in the same fashion: each rule has to be indexed over
arbitrary contexts and come equipped with a substitution rule. There
is no need to add more structural (substitution calculus) rules.

@@

Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
$\Pi\eta$ conditional. This needs another change: we substitute
$\Pi\beta$ for a more general variant which is also conditional on the
same equation.
\begin{alignat*}{10}
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
\end{alignat*}
But first we need to check that the usage of $\Pi\beta$ above when
proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
\begin{alignat*}{10}
  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
  & && \hat{B}[\langle a\rangle].
\end{alignat*}
Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
\begin{alignat*}{10}
  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
  & \El\,(\cd\,B) && {=}(\U\beta) \\
  & B
\end{alignat*}
This concludes the derivation of an even more minimalistic first-order
theory for the SOGAT (\ref{eq:tt}), relying essentially on the
universe and $\Pi$ types. It is clear from the construction that the
resulting theory is isomorphic to the one presented in Section
\ref{sec:tt} (and Appendix \ref{app:tt}), all equations are
interderivable. For reference, we list all the rules of the minimised
calculus in Appendix \ref{app:minimisation}.

@@

Rules of the theory derived in Section \ref{sec:minimisation}
(omitting $\Lift$):
\begin{alignat*}{10}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\N\ra\Set \\
  & \diamond && : \Con \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
  & \Sub && : \Con\ra\Con\ra\Set \\
  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & [\p][^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_e b[\gamma][\p] \\
  & \q[^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra \q[\gamma^+] =_e \q \\
  & [\p][\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra b[\p ][\langle a\rangle] =_e b \\
  & \q[\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra \q[\langle a\rangle] =_e a \\
  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
  & {\cdot}[]' && : (e:B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
  & && \hphantom{{}:{}} (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma]) \\
  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b \\
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
